---
title: "Merge Data Sets - Create Tidy Data"
author: "Baldvin Einarsson"
date: "March 12, 2016"
output: 
  html_document:
    keep_md: true
---

```{r setup, echo=FALSE, purl=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(knitr)
opts_knit$set(comment=NA)

# NB: We need to allow duplicate labels in order to purl
#  (Even though nothing is duplicated)
options(knitr.duplicate.label = 'allow')
purl(input="run_analysis.Rmd", output="run_analysis.R", documentation = 1)
```

We assume that the data has been downloade and unzipped, the location of which is the folder `UCI HAR Dataset/`:
```{r setFolderLoc}
dataLoc <- "UCI HAR Dataset/"
```

# Get Labels and Feature Names

First, we get the activity labels:
```{r getLabels}
activityLabels <- read.table(paste0(dataLoc,"activity_labels.txt"),
                             header = FALSE,
                             col.names = c("Value","Activity"),
                             stringsAsFactors = FALSE
                             )
```

Then, we obtain the names of the features:
```{r getFeatures}
features <- read.table(paste0(dataLoc,"features.txt"),
                       header=FALSE,
                       row.names = 1,
                       col.names = c("RowNum","Name"),
                       stringsAsFactors = FALSE)
```

Since we're only interested in the means and standard deviations, we find out which features have "mean" and "std" in their names:
```{r findMeanStdFeatures}
meanFeatures <- grepl("mean",features$Name)
stdFeatures <- grepl("std",features$Name)
```

# Read Data

First, we read the training set:
```{r getTrainingSet}
trainingData <- cbind(read.table(file=paste0(dataLoc,"train/subject_train.txt"),
                                 col.names = "SubjectID"),
                      read.table(file=paste0(dataLoc,"train/y_train.txt"), 
                                 col.names = "Activity"),
                      read.table(file=paste0(dataLoc,"train/X_train.txt"), 
                                 col.names = features$Name))
```
And then the test set:
```{r getTestSet}

testData <- cbind(read.table(file=paste0(dataLoc,"test/subject_test.txt"), 
                               col.names = "SubjectID"),
                  read.table(file=paste0(dataLoc,"test/y_test.txt"), 
                               col.names = "Activity"),
                  read.table(file=paste0(dataLoc,"test/X_test.txt"), 
                               col.names = features$Name))
```


# Merge

We now create a data frame from the above two types of data. However, we only extract the means and standard deviations of each measurement. Note that I have therefore skipped the columns which are the Fourier transformations of the measurements.

```{r mergeAndSubset}
selectedColumns <- c("SubjectID", "Activity",
                     grep("^[^f].+\\.mean\\.",names(testData),value = TRUE),
                     grep("^[^f].+std",names(testData),value = TRUE))

combinedData <- rbind(trainingData[,selectedColumns],
                      testData[,selectedColumns])

# Clear the raw data and flush memory
rm(list=c("trainingData","testData")); gc()
```

Ok, this leaves us with a total of `r nrow(combinedData)` rows and `r ncol(combinedData)` columns. Let's now rename the columns more clearly:

```{r renameColumns}
# First, we remove the "t" at the start of variable names:
names(combinedData) <- gsub("^t","",names(combinedData))
# Change "Acc" to a more descriptive "Acceleration":
names(combinedData) <- gsub("Acc","Acceleration",names(combinedData))
# Change "Mag" to a more descriptive "Magnitude":
names(combinedData) <- gsub("Mag","Magnitude",names(combinedData))
# Change "Gyro" to a more descriptive "Gyroskope":
names(combinedData) <- gsub("Gyro","Gyroskope",names(combinedData))
# Change "std" to "sd" (for R-style 'standard deviation'):
names(combinedData) <- gsub("\\.std","\\.sd",names(combinedData))


# Then, we clean the dots and such:
names(combinedData) <- gsub("\\.{2,3}","",names(combinedData))

```

Next, we add descriptive values to the activity:

```{r descriptiveActivity}
combinedData$Activity <- activityLabels$Activity[combinedData$Activity]
```


