---
title: "Merge Data Sets - Create Tidy Data"
author: "Baldvin Einarsson"
date: "March 12, 2016"
output: 
  html_document:
    keep_md: true
---

```{r setup, echo=FALSE, purl=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(knitr)
opts_knit$set(comment=NA)

# NB: We need to allow duplicate labels in order to purl
#  (Even though nothing is duplicated)
options(knitr.duplicate.label = 'allow')
purl(input="run_analysis.Rmd", output="run_analysis.R", documentation = 1)
```

We assume that the data has been downloade and unzipped, the location of which is the folder `UCI HAR Dataset/`:
```{r setFolderLoc}
dataLoc <- "UCI HAR Dataset/"
```

# Get Labels and Feature Names

First, we get the activity labels:
```{r getLabels}
activityLabels <- read.table(paste0(dataLoc,"activity_labels.txt"),
                             header = FALSE,
                             col.names = c("Value","Activity"),
                             stringsAsFactors = FALSE
                             )
```

Then, we obtain the names of the features:
```{r getFeatures}
features <- read.table(paste0(dataLoc,"features.txt"),
                       header=FALSE,
                       row.names = 1,
                       col.names = c("RowNum","Name"),
                       stringsAsFactors = FALSE)
```

Since we're only interested in the means and standard deviations, we find out which features have "mean" and "std" in their names:
```{r findMeanStdFeatures}
meanFeatures <- grepl("mean",features$Name)
stdFeatures <- grepl("std",features$Name)
```

# Read Data

First, we read the training set:
```{r getTrainingSet}
trainingSubject <- read.table(file=paste0(dataLoc,"train/subject_train.txt"), 
                               col.names = "SubjectID")
trainingActivity <- read.table(file=paste0(dataLoc,"train/y_train.txt"), 
                               col.names = "ActivityNumber")
trainingFeatures <- read.table(file=paste0(dataLoc,"train/X_train.txt"), 
                               col.names = features$Name)
```
And then the test set:
```{r getTestSet}
testSubject <- read.table(file=paste0(dataLoc,"test/subject_test.txt"), 
                               col.names = "SubjectID")
testActivity <- read.table(file=paste0(dataLoc,"test/y_test.txt"), 
                               col.names = "ActivityNumber")
testFeatures <- read.table(file=paste0(dataLoc,"test/X_test.txt"), 
                               col.names = features$Name)
```


# Merge

We now create a data frame from the above two types of data

```{r}
combinedData
```

